### 2024-08-10 machine-learning.qmd
20240810 - random forests

```{r}
library("LongituRF")
library("doParallel")
library("missForest")
library("powerjoin")
library("mice")
library("miceadds")
library("tidyverse")
```

@@

seasonalData <- powerjoin::power_full_join(
  playerSeasonMerged,
  playerMerged,
  by = "gsis_id",
  conflict = coalesce_xy
)

seasonalAndWeeklyData <- powerjoin::power_full_join(
  playerSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy
)

@@

seasonalData %>% 
  group_by(gsis_id, season) %>% 
  filter(n() > 1) %>% 
  head()
```

@@

```{r}
seasonalData %>% select_if(~class(.) == "Date")
seasonalData %>% select_if(is.character)
seasonalData %>% select_if(is.factor)
seasonalData %>% select_if(is.logical)

seasonalData_subset <- seasonalData %>% 
  dplyr::select(-any_of(c(
    "birth_date", "loaded", "full_name", "player_name", "player_display_name", "display_name", "suffix", "headshot_url", "player", "pos",
    "espn_id", "sportradar_id", "yahoo_id", "rotowire_id", "pff_id", "fantasy_data_id", "sleeper_id", "pfr_id",
    "pfr_player_id", "cfb_player_id", "pfr_player_name", "esb_id", "gsis_it_id", "smart_id",
    "college", "college_name", "team_abbr", "current_team_id", "college_conference", "draft_club", "status_description_abbr",
    "status_short_description", "short_name", "headshot", "uniform_number", "jersey_number", "first_name", "last_name",
    "football_name", "team"
  ))) %>% 
  filter(!is.na(season))
```

@@

```{r}
seasonalData_subset <- seasonalData_subset %>% mutate(across(where(is.character), as.factor))
seasonalData_subset <- seasonalData_subset %>% mutate(across(where(is.logical), as.factor))

seasonalData_subset %>% select_if(~class(.) == "Date") %>% str()
seasonalData_subset %>% select_if(is.character) %>% str()
seasonalData_subset %>% select_if(is.factor) %>% str()
seasonalData_subset %>% select_if(is.logical) %>% str()
```

```{r}
seasonalData_subset <- seasonalData_subset %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  mutate(
    fantasyPoints_lag = lead(fantasy_points)
  ) %>% 
  ungroup()
```

@@

- CONSIDER WIDENING THE DATA TO AVOID MULTILEVEL IMPUTATION
- IMPUTE THE TRAINING AND TEST DATA SEPARATELY

#### `missForest`

```{r}
num_cores <- detectCores()
num_cores

registerDoParallel(cores = num_cores)

seasonalData_subset_imp <- missForest::missForest(
  seasonalData_subset %>% select(ageCentered20, ageCentered20Quadratic, years_of_experience, position_group, fantasyPoints_lag) %>% data.frame,
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")
```

#### `MICE`

```{r}
# Specify data to impute
dataToImpute <- seasonalData_subset %>% 
  select(gsis_id, season, ageCentered20, ageCentered20Quadratic, fantasy_points, years_of_experience) %>% 
  filter(!is.na(gsis_id), !is.na(season)) %>% 
  mutate(ID = as.integer(as.factor(as.character(gsis_id)))) %>% 
  arrange(gsis_id, season)

# Specify variables to impute
varsToImpute <- c("fantasy_points")
Y <- varsToImpute

# Specify predictors of missing data
clusterVariable <- "ID"
fixedEffectPredictors <- c("ageCentered20Quadratic")
randomEffectPrectors <- c("ageCentered20")

# Specify number of imputations
numImputations <- 1

# Specify the imputation method
meth <- make.method(dataToImpute)
meth[1:length(meth)] <- ""
meth[Y] <- "2l.pmm" # specify the imputation method here; this can differ by outcome variable

# Specify the predictor matrix

pred <- make.predictorMatrix(dataToImpute)
pred[1:nrow(pred), 1:ncol(pred)] <- 0
pred[Y, clusterVariable] <- (-2) # cluster variable
pred[Y, randomEffectPrectors] <- 2 # random effect predictor
#pred[Y, fixedEffectPredictors] <- 1 # fixed effect predictor
pred[Y, Y] <- 1 # fixed effect predictor of each predictor in predicting the other predictors

diag(pred) <- 0
pred

# Imputation
mi_mice <- mice(
  as.data.frame(dataToImpute),
  method = meth,
  predictorMatrix = pred,
  m = numImputations,
  maxit = 100, # generally use 100 maximum iterations; this example uses 5 for speed
  seed = 52242)

mi_mice$loggedEvents

seasonalData_subset_imp <- complete(
  mi_mice,
  action = "stacked",
  include = FALSE)
```

@@

@@

```{r}
smerf <- LongituRF::MERF(
  X = seasonalData_subset_imp$ximp[,c("position_group")] %>% as.matrix(),
  Y = seasonalData_subset_imp$ximp[,c("fantasyPoints_lag")],
  Z = seasonalData_subset_imp$ximp[,c("years_of_experience")] %>% as.matrix(),
  id = seasonalData_subset$gsis_id,
  time = seasonalData_subset_imp$ximp[,c("ageCentered20")] %>% as.matrix(),
  mtry = 2,
  ntree = 500,
  sto = "BM")

smerf$forest # is the fitted random forest (obtained at the last iteration)
smerf$random_effects # are the predicted random effects for each individual
smerf$omega # are the predicted stochastic processes
plot(smerf$Vraisemblance) # evolution of the log-likelihood
smerf$OOB # OOB error at each iteration
```

>>>>

```{r}
library("LongituRF")
library("doParallel")
library("missForest")
library("powerjoin")
library("mice")
library("miceadds")
library("caret")
library("tidyverse")
```

@@

seasonalData <- powerjoin::power_full_join(
  playerSeasonMerged,
  playerMerged,
  by = "gsis_id",
  conflict = coalesce_xy
) %>% 
  filter(!is.na(season)) %>% 
  select(gsis_id, season, player_display_name, position, recent_team, games, everything()) #-todo: deduplicate team from prior merged objects

seasonalAndWeeklyData <- powerjoin::power_full_join(
  playerSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy
) %>% 
  filter(!is.na(week))

@@

seasonalData %>% 
  group_by(gsis_id, season) %>% 
  filter(n() > 1) %>% 
  head()

seasonalAndWeeklyData %>% 
  group_by(gsis_id, season, week) %>% 
  filter(n() > 1) %>% 
  head()
```

### Fill in Missing Data {#sec-fillMissingData}

```{r}
seasonalData <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  fill(
    player_name, player_display_name, pos, position, position_group,
    .direction = "downup") %>% 
  ungroup()
```

```{r}
seasonalData_lag <- seasonalData %>% 
  arrange(gsis_id, season) %>% 
  group_by(gsis_id) %>% 
  mutate(
    fantasyPoints_lag = lead(fantasy_points)
  ) %>% 
  ungroup()

seasonalData_lag %>% 
  select(gsis_id, player_display_name, season, fantasy_points, fantasyPoints_lag)
```

### Create New Data Object {#sec-newDataObject}

```{r}
newData_seasonal <- seasonalData_lag %>% 
  filter(season == max(season, na.rm = TRUE))

# Drop rows with missing fantasyPoints_lag (i.e., the player's last season), so we don't impute criterion (fantasy points)
seasonalData_lag <- seasonalData_lag %>% 
  filter(!is.na(fantasyPoints_lag))
```

@@

```{r}
seasonalData_lag %>% select_if(~class(.) == "Date")
seasonalData_lag %>% select_if(is.character)
seasonalData_lag %>% select_if(is.factor)
seasonalData_lag %>% select_if(is.logical)

seasonalData_lag_subset <- seasonalData_lag %>% 
  dplyr::select(-any_of(c(
    "birth_date", "loaded", "full_name", "player_name", "player_display_name", "display_name", "suffix", "headshot_url", "player", "pos",
    "espn_id", "sportradar_id", "yahoo_id", "rotowire_id", "pff_id", "fantasy_data_id", "sleeper_id", "pfr_id",
    "pfr_player_id", "cfb_player_id", "pfr_player_name", "esb_id", "gsis_it_id", "smart_id",
    "college", "college_name", "team_abbr", "current_team_id", "college_conference", "draft_club", "status_description_abbr",
    "status_short_description", "short_name", "headshot", "uniform_number", "jersey_number", "first_name", "last_name",
    "football_name", "team"
  )))
```

@@

```{r}
seasonalData_lag_subset <- seasonalData_lag_subset %>% mutate(across(where(is.character), as.factor))
seasonalData_lag_subset <- seasonalData_lag_subset %>% mutate(across(where(is.logical), as.factor))

seasonalData_lag_subset %>% select_if(~class(.) == "Date") %>% str()
seasonalData_lag_subset %>% select_if(is.character) %>% str()
seasonalData_lag_subset %>% select_if(is.factor) %>% str()
seasonalData_lag_subset %>% select_if(is.logical) %>% str()
```

### Separate by Position {#sec-separateByPosition}

```{r}
# Training Data
seasonalData_lag_subsetQB <- seasonalData_lag_subset %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    completions:rushing_2pt_conversions, contains(".pass"))

seasonalData_lag_subsetRB <- seasonalData_lag_subset %>% 
  filter(position == "RB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))

seasonalData_lag_subsetWR <- seasonalData_lag_subset %>% 
  filter(position == "WR") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))

seasonalData_lag_subsetTE <- seasonalData_lag_subset %>% 
  filter(position == "TE") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))

# Out-of-Sample Prediction
newData_seasonalQB <- newData_seasonal %>% 
  filter(position == "QB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    completions:rushing_2pt_conversions, contains(".pass"))

newData_seasonalRB <- newData_seasonal %>% 
  filter(position == "RB") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))

newData_seasonalWR <- newData_seasonal %>% 
  filter(position == "WR") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))

newData_seasonalTE <- newData_seasonal %>% 
  filter(position == "TE") %>% 
  select(
    gsis_id, season, games, gs, years_of_experience, age, ageCentered20, ageCentered20Quadratic,
    height, weight, rookie_year, draft_number,
    fantasy_points, fantasy_points_ppr, fantasyPoints_lag,
    carries:special_teams_tds, contains(".rush"), contains(".rec"))
```

@@

- SEPARATE BY POSITION
- CONSIDER WIDENING THE DATA TO AVOID MULTILEVEL IMPUTATION
- IMPUTE THE TRAINING AND TEST DATA SEPARATELY

Here is a vignette demonstrating how to impute missing data using `missForest()`: <https://rpubs.com/lmorgan95/MissForest> (archived at: <https://perma.cc/6GB4-2E22>).

```{r}
#| eval: false

num_cores <- detectCores() - 1
```

```{r}
#| include: false

num_cores <- detectCores()
```

```{r}
num_cores

cl <- makeCluster(num_cores)
registerDoParallel(cl)

set.seed(52242)

# QBs
seasonalData_lag_subsetQB_imp <- missForest::missForest(
  seasonalData_lag_subsetQB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetQB_imp$OOBerror

newData_seasonalQB_imp <- missForest::missForest(
  newData_seasonalQB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalQB_imp$OOBerror

# RBs
seasonalData_lag_subsetRB_imp <- missForest::missForest(
  seasonalData_lag_subsetRB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetRB_imp$OOBerror

newData_seasonalRB_imp <- missForest::missForest(
  newData_seasonalRB %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalRB_imp$OOBerror

# WRs
seasonalData_lag_subsetWR_imp <- missForest::missForest(
  seasonalData_lag_subsetWR %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetWR_imp$OOBerror

newData_seasonalWR_imp <- missForest::missForest(
  newData_seasonalWR %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalWR_imp$OOBerror

# TEs
seasonalData_lag_subsetTE_imp <- missForest::missForest(
  seasonalData_lag_subsetTE %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

seasonalData_lag_subsetTE_imp$OOBerror

newData_seasonalTE_imp <- missForest::missForest(
  newData_seasonalTE %>% 
    select(-gsis_id) %>%
    data.frame(),
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")

newData_seasonalTE_imp$OOBerror

stopCluster(cl)
```

@@

```{r}
cl <- makeCluster(num_cores)
registerDoParallel(cl)

set.seed(52242)

randomForest_qb <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_subsetQB_imp$ximp,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_rb <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_subsetRB_imp$ximp,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_wr <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_subsetWR_imp$ximp,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

randomForest_te <- caret::train(
  fantasyPoints_lag ~ ., # use all predictors
  data = seasonalData_lag_subsetTE_imp$ximp,
  method = "rf",
  trControl = trainControl(
    method = "cv",
    number = 10)) # 10-fold cross-validation

stopCluster(cl)

print(randomForest_qb)
print(randomForest_rb)
print(randomForest_wr)
print(randomForest_te)

newData_seasonalQB_imp$ximp$fantasyPoints_lag <- predict(
  randomForest_qb,
  newdata = newData_seasonalQB_imp$ximp
)

newData_seasonalRB_imp$ximp$fantasyPoints_lag <- predict(
  randomForest_rb,
  newdata = newData_seasonalRB_imp$ximp
)

newData_seasonalWR_imp$ximp$fantasyPoints_lag <- predict(
  randomForest_wr,
  newdata = newData_seasonalWR_imp$ximp
)

newData_seasonalTE_imp$ximp$fantasyPoints_lag <- predict(
  randomForest_te,
  newdata = newData_seasonalTE_imp$ximp
)

newData_seasonalQB$fantasyPoints_lag <- newData_seasonalQB_imp$ximp$fantasyPoints_lag
newData_seasonalRB$fantasyPoints_lag <- newData_seasonalRB_imp$ximp$fantasyPoints_lag
newData_seasonalWR$fantasyPoints_lag <- newData_seasonalWR_imp$ximp$fantasyPoints_lag
newData_seasonalTE$fantasyPoints_lag <- newData_seasonalTE_imp$ximp$fantasyPoints_lag

newData_seasonalQB <- left_join(
  newData_seasonalQB,
  newData_seasonal %>% select(gsis_id, player_display_name, recent_team),
  by = "gsis_id"
)

newData_seasonalRB <- left_join(
  newData_seasonalRB,
  newData_seasonal %>% select(gsis_id, player_display_name, recent_team),
  by = "gsis_id"
)

newData_seasonalWR <- left_join(
  newData_seasonalWR,
  newData_seasonal %>% select(gsis_id, player_display_name, recent_team),
  by = "gsis_id"
)

newData_seasonalTE <- left_join(
  newData_seasonalTE,
  newData_seasonal %>% select(gsis_id, player_display_name, recent_team),
  by = "gsis_id"
)

newData_seasonalQB %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_seasonalRB %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_seasonalWR %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)

newData_seasonalTE %>%
  arrange(-fantasyPoints_lag) %>% 
  select(gsis_id, player_display_name, fantasyPoints_lag)
```

@@

```{r}
#| eval: false

smerf <- LongituRF::MERF(
  X = seasonalData_subsetQB_imp$ximp[,c("passing_epa")] %>% as.matrix(),
  Y = seasonalData_subsetQB$fantasyPoints_lag,
  Z = seasonalData_subsetQB_imp$ximp[,c("pacr")] %>% as.matrix(),
  id = seasonalData_subsetQB$gsis_id,
  time = seasonalData_subsetQB_imp$ximp[,c("ageCentered20")] %>% as.matrix(),
  ntree = 500,
  sto = "BM")

smerf$forest # the fitted random forest (obtained at the last iteration)
smerf$random_effects # the predicted random effects for each player
smerf$omega # the predicted stochastic processes
plot(smerf$Vraisemblance) # evolution of the log-likelihood
smerf$OOB # OOB error at each iteration
```
